---
title: "GSPP R Tutorial Day 2"
author: "Chris Kennedy"
date: "February 2, 2016"
output: html_document
---

Start with an exercise from day 1

[Review function vs vector vs. element vs. characters ]

[Use dataset that is more fun, like the imdb dataset.]

# Load a new dataset
```{r}
library(readr)
# Via https://github.com/fivethirtyeight/data/blob/master/bad-drivers/bad-drivers.csv
# https://raw.githubusercontent.com/fivethirtyeight/data/master/bad-drivers/bad-drivers.csv
data = read_csv("http://is.gd/gspp_2016")
names(data)
names(data) <- c("state", "fatal", "speeding", "alcohol",
                 "not_distracted", "no_prev", "premiums", "losses")
str(data)
```

**Class Practice**

* How many rows and columns in the dataset?
* What is the variable type for each column?
* What do the distributions look like?
* Can you plot histograms for the continuous variables?
* How many observations per state are there?

```{r}
table(data$state)
```

# missing values - is.na, na.rm
```{r}
is.na(c(1, 2, 3))
is.na(c(1, NA, 3))

is.na(data$alcohol)
```
** How do we count how many of those are TRUEs?

```{r}
sum(is.na(data[, 1]))
sum(is.na(data[, "fatal"]))
sum(is.na(data[, "speeding"]))
```

Let's set a few values to missing for all columns except state.

```{r}
data[c(3, 7, 17, 37), -1] <- NA
```

Now do we find missing values?

```{r}
apply(data, MARGIN=2, FUN=function(x){ sum(is.na(x)) })
```

With missing data, what's the mean fatality rate?
```{r}
mean(data$fatal)
mean(data$fatal, na.rm=T)
```

For safety, R returns an NA for mathematical functions when it is passed in data with missing values. The `na.rm` option tells R that it can omit the NAs and give us the result.

# More data frames

Adding a column to our dataframe.
```{r}
set.seed(1)
rand = rnorm(nrow(data))
length(rand)

data = cbind(data, rand)
dim(data)
names(data)
data$rand
```

Creating a dataframe from two individual vectors.

```{r}
state = data$state
dim(state)
nrow(state)
length(state)
class(state)

premiums = data$premiums

# Not right - do you remember why?
state_and_premiums = data.frame(cbind(data$state, data$premiums))
str(state_and_premiums)

# Better:
state_and_premiums = data.frame(cbind(data$state, data$premiums))
class(state_and_premiums)
```

```{r}
summary(data)
# Could show one version without using na.rm=T
guam = c("Guam", mean(data$fatal), mean(data$speeding), mean(data$alcohol), mean(data$not_distracted), mean(data$no_prev), mean(data$premium), mean(data$losses), rnorm(1))
guam
length(guam)
dim(data)

# Remove NAs
guam = c("Guam", mean(data$fatal, na.rm=T), mean(data$speeding, na.rm=T), mean(data$alcohol, na.rm=T), mean(data$not_distracted, na.rm=T), mean(data$no_prev, na.rm=T), mean(data$premium, na.rm=T), mean(data$losses, na.rm=T), rnorm(1))
guam

new_df = rbind(data, guam)
new_df[new_df$state == "Guam", ]
subset(new_df, state == "Guam")
```

A handy function is colMeans to make this easier.

```{r error=T}

# This doesn't work.
colMeans(data)

# This doesn't work either!
colMeans(data, na.rm=T)

# What about this?
means = colMeans(data[, -1], na.rm=T)

# Update the random number.
means[8] = rnorm(1)

puerto_rico = c("Puerto Rico", means)

new_df2 = rbind(data, guam, puerto_rico)
new_df2[50:53, 1:3]
```

rowMeans is similar, although generally not as useful.
  
#- aggregate - come back to if there's time.
  
# more ggplot2
```{r}
library(ggplot2)
# Histogram - with one variable default geom = "histogram"
qplot(premium, data=data)

# Scatter plot - because we have two variables, geom = "scatterplot".
qplot(premium, losses, data=data)
qplot(premium, losses, data=data, xlab="premiums", ylab="losses", main="state insurance losses vs premiums")

# Fix the horrible gray default.
qplot(premium, losses, data=data, xlab="premiums", ylab="losses", main="state insurance losses vs premiums") + theme_bw()

# Change color - I() because it is "aesthetics"
qplot(premium, losses, color=I("blue"), data=data, xlab="premiums", ylab="losses", main="state insurance losses vs premiums") + theme_bw()

# Color by a third variable
qplot(premium, losses, color=fatal, data=data, xlab="premiums", ylab="losses", main="state insurance losses vs premiums") + theme_bw()

# Change dot size.
qplot(premium, losses, color=fatal, data=data, xlab="premiums", ylab="losses", main="state insurance losses vs premiums", size=I(5)) + theme_bw()

# Change dot size by a 4th variable.
qplot(premium, losses, color=fatal, data=data, xlab="premiums", ylab="losses", main="state insurance losses vs premiums", size=speeding) + theme_bw()

# Redo without using qplot.
ggp = ggplot(data=data, mapping=aes(x=premium, y=losses, color=fatal, size=speeding))

ggp

# We need to specify the geometry, or plot type.
ggp + geom_point()
ggp + geom_point() + theme_bw()

# Density plot
ggplot(data=data, aes(x=premium)) + geom_density()
# Compare to histogram
qplot(data$premium)
# Change the color
ggplot(data=data, aes(x=premium)) + geom_density(color="red") + theme_bw()

# Abstract art.
ggp + geom_path() + theme_bw()

# NOICE - ordered by the x-axis.
ggp + geom_line() + theme_bw()

# Let's go 80s arcade style.
ggp + geom_step() + theme_bw()

# Change geometry to the labels themselves.
# Note that we could have specified the aes(label=state) in the original ggplot.
textplot <- ggp + labs(x="premiums", y="losses") + 
  ggtitle("state insurance losses vs premiums") + 
  geom_text(mapping=aes(label=state), data=data) + theme_bw()
# Could do: geom_text(mapping=aes(premium, losses, label=state), data=data)
textplot

# Hide labels that overlap.
textplot + geom_text(mapping=aes(label=state), data=data, check_overlap=T)

# Labels with a background; no overlap checking though :/
textplot + geom_label(mapping=aes(label=state), data=data)

# Point and text, with text offset
# vjust is relative to the point itself. 
# Nudge moves the point up from its default location.
ggp + geom_point(size=3) + geom_text(mapping=aes(label=state), data=data, vjust="bottom", nudge_y=1.5, size=3)

# Scale the color of the text.
textplot + scale_color_gradient(low="blue", high="gold") # different colors

# Change geometry to the labels themselves.
textplot <- ggp + labs(x="premiums", y="losses") + 
  ggtitle("state insurance losses vs premiums") + 
  geom_text(mapping=aes(label=state), data=data) + theme_bw()
# Could do: geom_text(mapping=aes(premium, losses, label=state), data=data)
textplot

# Export a plot.
ggsave(filename="mychart.pdf", plot=ggp, scale=, width=, height=)
ggsave(filename="mychart.png", plot=ggp, scale=, width=, height=)

```

**Ggplot resources **

# Flow control

**If statements**

```{r}
if (T) {
  cat("Help, I'm being executed!\n")
}

if (F) {
  cat("Yay, I'm not executed!")
}

if (8 > 3 & 2 < 5) {
  cat("Do whatever I want.\n")
}

# T-statistic example
set.seed(1)
t = rnorm(1)
t
if (t > 0) {
  cat("t is a positive pedro.")
} else if (t < 0) {
  cat("t is a negative nancy")
  # Set to be positive.
  t = t * -1
} else {
  cat("t is exactly zero, weird!")  
}
```

** For loops **

```{r}
for (i in 1:10) {
  cat(i, " ")
}

# Print is overly verbose.
for (i in 1:10) {
  print(i)
}

# Looping over the rows of a dataframe.
for (row in 1:nrow(data)) {
  cat("\nWe are on row:", row, "\n")
  # Print the first element of each row.
  print(data[row, 1])
  observation = data[row, ]
  # Do something with the observation.
}

```

**Class**: Can you loop over the columns of a dataframe, and print out how many NAs are in each column?
  
# Functions
```{r error=T}
blah = function() {
  answer = 5 + 3
  cat("The answer is:\n")
  print(answer)
}
blah()

blah = function(x) {
  answer = x + 3
  cat("The answer is:\n")
  print(answer)
}
blah()
blah(5)

blah = function(x = 5) {
  answer = x + 3
  cat("The answer is:\n")
  print(answer)
}
blah
blah()
blah(10)
blah(x = 10)
# What will happen if we do blah(x <- 10)?

```

Let's return some data
```{r}

another = function(my_var) {
  answer = sum(my_var^2) - 10  
  return(answer)
}
another(10)
another(10:11)
result = another(10:11)
result

another = function(my_var) {
  answer = sum(my_var^2) - 10  
  cat("The answer is\n")
  print(answer)
  return(answer)
}
another(6)
# Why did it print out twice?
result = another(6)

another = function(my_var, print = T) {
  answer = sum(my_var^2) - 10  
  if (print) {
    cat("The answer is\n")
    print(answer)
  }
  return(answer)
}
another(5)
another(6, print=F)
another(6, F)
result = another(6, F)

last_one = function() {
  t_stat = 5.5
  p_value = 0.001
  result = c(t_stat, p_value)
  return(result)
}

last_one()
answer = last_one()
answer
answer[1]
answer[2]

last_one_better = function() {
  t_stat = 5.5
  p_value = 0.001
  result = c(t=t_stat, p=p_value)
  return(result)
}
answer = last_one_better()
answer
answer[1]
answer["t"]
answer["p"]

```

# more working with vectors

```{r}
# Let's go back to the transportation data.
summary(data)

# Ifelse
above_average = ifelse(data$fatal >= mean(data$fatal, na.rm=T), 1, 0)
table(above_average)

# %in%
data$state %in% c("California", "District of Columbia", "Texas")
sum(data$state %in% c("California", "District of Columbia", "Texas", "nuthin"))

# Ifelse with %in%
good_states = ifelse(data$state %in% c("California", "District of Columbia", "Texas"), 1, 0)
table(good_states)

# Exclude certain columns by name
ncol(data)

names(data)
# Same thing
colnames(data)

names(data) %in% c("no_prev", "premium", "rand")
!names(data) %in% c("no_prev", "premium", "rand")
subset = data[, !names(data) %in% c("no_prev", "premium", "rand")]
names(subset)

```

**Paste and paste0**

These are used for output primarily, combining vectors into printable strings.

```{r}
vec = c(10, 5, 9, 6, 13)
paste(vec, "is a number")
# Convert to a single string, with each element separated by a comma.
paste(vec, collapse=", ")
cat("My output is:", vec)
cat("My output is:", paste(vec, collapse=", "))

cat(paste(vec))
cat(paste0(vec))
```

# Other topics if we have time:  
* apply and tapply (skip possibly)
* sample
* merging files
* regression
  - OLS
  - logistic
  - predict()
  - interaction terms
* few last things
  - system.time()
  - gc()
