---
title: "R Training Day 1"
author: "Chris Kennedy"
date: "8/25/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## New headline

Checking about new changes.

```{r}
2 + 3
1:3
c("first element", "2nd element")

3 * 3 * 3 * 3 * 3
3^7
sqrt(16)
log(10)
exp(2.3)
mean(10)
mean(10, 11, 12)
mean(c(10, 11, 12))
my_numbers <- c(10, 11, 12)
my_numbers2 = c(10, 11, 12)
my_numbers
```
Now I'm going to print my_numbers.
```{r error=TRUE}
# Display the values of the my_numbers vector.
print(my_numbers)
# This outputs my_numbers2.
print(my_numbers2) 

another.var <- c(5, 7, 9, 11)
another.var
(another.var <- c(5, 7, 9, 11))

(blah = "program evaluation")

?c
c(1, 7:9)
c(1:5, 10.5, "next")
c(1, 2, c("next", "one"))

var = c(1:5, 10.5, "next")
class(var)
class(1:5)
?class

ls()

class(var)
var = c(1:5, 10.5, "next")

rm(my_numbers, my_numbers2)
ls()

never_made_this_one

install.packages(rio)
install.packages("rio")
library(rio)
install_formats()

help(package = rio)
print("test")

data = import("https://stats.idre.ucla.edu/stat/stata/dae/binary.dta")
# Typo gives a slightly unhelpful error:
data = import("https://stats.idre.ucla.edu/stat/stata/dae/biny.dta")
```
## Break time until 1:50pm

```{r error=TRUE}
names(data)
colnames(data)
dim(data)
nrow(data)
ncol(data)
summary(data)
# sum() is used to calculate the sum of a vector:
# (unlike Stata)
sum(c(10, 5, 1))
str(data)
head(data)
?head
head(data, 2)
head(x = data, n = 2)
head(data, n = 10)
data
data[]
data[15, ]
data[15, "gpa"]
data[15, "gpa", "admit"]
data[15, c("gpa", "admit")]
columns_to_pick = c("gpa", "admit")
data[15, columns_to_pick]
drop_f = data[15, "gpa", drop = F]
drop_f
class(drop_f)
drop_t = data[15, "gpa", drop = T]
drop_t
class(drop_t)
data[15, "gpa", drop = TRUE]
data[, "blah"]

tail(data)
data[, "gre"]
data[, c("gre")]
data$gre
mean(data$gre)
summary(data$gre)
sd(data$gre)
hist(data$gre)

install.packages("ggplot2")
library(ggplot2)
qplot(data$gre)
qplot(data$gre) + theme_bw()
qplot(data$gre, data$gpa)

table(data$admit)

data$admit == 1
table(data$admit == 1)
table(data$admit == 1, data$admit)

data[data$admit == 1,]
these_rows = data$admit == 1
these_rows
# Same thing:
class(data)
class(these_rows)
data
length(these_rows)
data[these_rows, ]
# People with admit = 1, multiply gre score by 10
# and save the result.
data
data$new_gre = data$gre
ls()
names(data)
cor(data$gre, data$new_gre)
cor(data)

mean(data[data$admit == 1, "gre"])
hist(data[data$admit == 1, "gre"])
qplot(data[data$admit == 1, "gre"])

# Same result, here we subsetted a vector that we got from data$gre.
qplot(data$gre[data$admit == 1])

data[data$admit == 1, "new_gre"] = data[data$admit == 1, "new_gre"] * 10
summary(data[data$admit == 1, ])
summary(data[data$admit == 0, ])

```

This is normal text.

## What are these weird characters?

* Parentheses () - these can group parts of code, and are used after a function name to run that function. We pass in arguments inside of those parentheses. Eg: `r mean(c(5, 10))`. (Also example of inline R code.)
* Brackets [] - subsetting dataframes or vectors. Dataframe needs two arguments: first argument is which rows you want, second is which columns. A vector has only one argument for which elements you want.
* Dollar sign $ - shortcut to choose one variable (column) from your dataframe and get its values as a vector.

```{r}
# Examples of using backticks to get help on operators:
?`$`
?`(`
```
