---
title: "R Workshop Day 2"
author: "Chris Kennedy"
date: "February 5, 2016"
output: html_document
---
Function - collection of commands that may return something, and may accept argument.

Data structures:
vector,
element,
character string,
object/variable
```{r}
1
c(1, 2, 3)
"my name"
c("str1", "str2")
c("str1", "str2", 5)

my_variable = 10

#install.packages("readr")
library(readr)

data = read_csv("http://is.gd/gspp_2016")
```
* How many rows and columns in the dataset?
- ncol, nrow, dim

* What is the variable type for each column?
- str(), 

```{r}
str(data)
structure(data)
names(data)
data[, "State"]
data[, 2]
data[, c(2, 5, 7)]
names(data)

new_names = c("state", "fatal", "speeding","not_distracted", "no_prev", "premium", "losses")
length(new_names)
length(names(data))
ncol(data)

names(data) <- c("state", "fatal", "speeding","alcohol", "not_distracted", "no_prev", "premium", "losses")
names(data)
```
* What do the distributions look like?

```{r}
summary(data)
```
* Can you plot histograms for the continuous variables?
```{r}
library(ggplot2)
qplot(data$fatal)
qplot(data[, 2])
qplot(data[, "fatal"])
which_column = "fatal"
data[, which_column]
qplot(data[, which_column])
# This doesn't work:
data$which_column
# This does:
data$fatal

```
* How many observations per state are there?
```{r}
table(data$state)
```

Missing values

TODO: lookup code for prettier tables.

```{r}
is.na
?is.na

is.na(5)
is.na(NA)

is.na("STRING")
is.na(c(1, 2, NA))
my_vec = c(1, 2, NA)
is.na(my_vec[1])
is.na(my_vec[2])
is.na(my_vec[3])
is.na(my_vec)

names(data)
is.na(data$speeding)

data$speding
is.na(NULL)
is.na(data$speding)

sum(is.na(data$speeding))

my_names = names(data)
my_names

length(my_names)
ncol(data)
1:length(my_names)

# Long way
for (current_column_number in 1:length(my_names)) {
  print(current_column_number)
}

# Short way
for (col_num in 1:ncol(data) ) {
  print(col_num)
  4 + 5
  blah = rnorm(10)
  answer = mean(blah) 
  print(answer)
}

for (x in 1:8) { print(x) }

5
qplot(data[, 2])
print(qplot(data[, 2]))

for (col_num in 1:ncol(data) ) {
  col_num
  print(col_num)
  #print(qplot(data[, col_num]))
  #print(qplot(data[, col_num]))
}

col_num
names(data)
for (col_num in 1:ncol(data) ) {
  print(col_num)
  #names_vector = names(data)
  #print(names_vector[1])
  #print(names_vector[col_num])
  variable_name <- names(data)[col_num]
  print(qplot(data[, col_num], xlab = variable_name))
  # Same thing:
  print(qplot(data[, col_num], xlab = names(data)[col_num]))
}

# Try to do a for loop to count the number of NAs in each column
for (col_num in 1:ncol(data)) {
  cat("Variable name: ")
  cat(names(data)[col_num])
  cat("\n")
  cat("Sum of missing values: \n")
  print(sum(is.na(data[, col_num])))

  cat("Sum of missing values:", sum(is.na(data[, col_num])), "\n") 
   
  cat("Table of missing values: \n")
  print(table(is.na(data[, col_num])))
}

data[c(3, 7, 17, 37), -1]

data[c(3, 7, 17, 37), -1] <- NA

# Don't have to do this:
data[c(3, 7, 17, 37), -1] <- c(NA, NA, NA, NA, ...)

data[, -c(1, 2, 3)]

# Run for loop again and count NAs in each column

# Single column:
sum(is.na(data[, 2]))

mean(data$fatal)

mean(data$fatal, na.rm = TRUE)

mean(data$fatal[!is.na(data$fatal)] )

!is.na(data$fatal)

data$fatal[!is.na(data$fatal)]

```

```{r}
set.seed(1)
rand <- rnorm(nrow(data))
length(rand)
rand
mean(rand)

set.seed(2)
rand <- rnorm(nrow(data))
mean(rand)
rand

set.seed(10)
rand <- rnorm(nrow(data))
mean(rand)

cbind

# Adding a column to a dataframe, create a new dataframe.
data2 <- cbind(data, rand)
names(data2)
data2$rand

# Overwrite the existing dataframe
data <- cbind(data, rand)
names(data)

names(data2)

# Dropping a column
data2 = data2[, -9]
names(data2)

# another way - delete this column.
data2$rand = NULL
names(data2)

id = 1:100
rand = rnorm(100)
my_dataframe = data.frame(cbind(id, rand))
class(my_dataframe)
dim(my_dataframe)

names(data2)
guam = c("Guam", mean(data$fatal), mean(data$speeding), mean(data$alcohol), 1, 1, 1, 1)
length(guam)
guam

data3 = rbind(data, guam)
nrow(data3)
data3[52, ]

data3 = data3[-52, ]
nrow(data3)

data3 = data3[1:51, ]

View(data3[, -1])

means <- colMeans(data[, -1])
means
length(means)

puerto_rico = c("Puerto Rico", means)
puerto_rico

another_df = rbind(data, puerto_rico)
View(another_df)
another_df[another_df$state == "Puerto Rico", ]
names(another_df)
summary(another_df$fatal)
class(another_df$fatal)
another_df$fatal < 10
another_df[another_df$fatal < 10, ]

puerto_rico = c(NA, means)
puerto_rico

another_df = rbind(data, puerto_rico)
another_df[52, ]
another_df$state[52] = "Puerto Rico"
another_df[52, ]
class(another_df$fatal)

another_df[another_df$fatal < 10, ]

```

```{r}
library(ggplot2)

qplot(data$premium)
# Can't do this:
qplot(premium)
# second data is referring to the dataframe.
# first data
qplot(premium, data = data)

qplot(premium, losses, data = data)

cor(data$premium, data$losses)

# Remove obs with NA in premium or losses.
data_no_nas = data[!is.na(data$premium) & !is.na(data$losses), ]
cor(data_no_nas$premium, data_no_nas$losses)

qplot(premium, losses, data = data, xlab="Premiums", ylab="Losses in $MM", main="State insurrance losses vs. premiums")

qplot(premium, losses, data = data, xlab="Premiums", ylab="Losses in $MM", main="State insurrance losses vs. premiums") + theme_bw()

qplot(premium, losses, data = data, xlab="Premiums", ylab="Losses in $MM", main="State insurrance losses vs. premiums", color=I("red")) + theme_bw()

qplot(premium, losses, data = data, xlab="Premiums", ylab="Losses in $MM", main="State insurrance losses vs. premiums", color=fatal) + theme_bw()

qplot(premium, losses, data = data, xlab="Premiums", ylab="Losses in $MM", main="State insurrance losses vs. premiums", color=fatal, size=I(5)) + theme_bw()

qplot(premium, losses, data = data, xlab="Premiums", ylab="Losses in $MM", main="State insurrance losses vs. premiums", color=fatal, size=speeding) + theme_bw()

ggp <- ggplot(data=data, mapping=aes(x=premium, y=losses, color=fatal, size=speeding)) + theme_bw()
ggp

ggp + geom_point()
my_plot = ggp + geom_point()
my_plot

qplot(data$premium)

ggplot(data=data, mapping=aes(x=premium)) + geom_histogram(aes(y=..density..)) + geom_density(color="red")

ggp + geom_path()
ggp + geom_line() + ggtitle("New title")

```

[Ran out of time for the remainder, but it's on github!]